---
title: "Analysis of Boston data"
author: "Jussi Jaatinen"
output: html_document
fig_caption: true
fig_width: 15
fig_height: 8
code_folding: hide
---

# Analysis of Boston data

### Overview of data

Boston data is included to R-package as a demonstartion or example.

Dataset contains information about Boston area. It includes following variables:

- crim = per capita crime rate by town             
- zn = proportion of residential land zoned for lots over 25,000 sq.ft.    
- indus =  proportion of non-retail business acres per town				
- chas = Charles River dummy variable 
- nox =  nitrogen oxides concentration (parts per 10 million)
- rm = average number of rooms per dwelling
- age  = proportion of owner-occupied units built prior to 1940
- dis = weighted mean of distances to five Boston employment centres
- rad = index of accessibility to radial highways
- tax =  full-value property-tax rate per $10000  
- ptratio = pupil-teacher ratio by town
- black = 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
- lstat = lower status of the population (percent)
- medv = median value of owner-occupied homes in $1000s

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Loading necessary libraries
library(MASS)
library(dplyr)
library(plotly)
library(corrplot)
library(ggplot2)
library(GGally)
# load the data
data("Boston")
```

##### Structure and the dimensions of the data 

Dataset has 14 variables and 506 observationsand all variables are numerical.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
str(Boston)
dim(Boston)

```

##### Summary, graphical presentation of data and correlations

As seen in pairs plot, most of the variables are not normally distributed. Most of them are skewed and some of them are bimodal. Correlations between variables are better viewed in correlation plotting, where on the upper-right side the biggest circles indicate highest correlations (blue = positive or red = negative). Corresponding number values are mirrored on the lower-left side.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
summary(Boston)
# pairs(Boston)
# ggpairs(Boston, mapping = aes(), lower = list(combo = wrap("facethist", bins = 40)), upper = list(continuous = wrap("cor", size = 12)))
ggpairs(Boston, mapping = aes(corSize = 8), lower = list(combo = wrap("facethist", bins = 100)))

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits = 2)
# visualize the correlation matrix
# corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
corrplot.mixed(cor_matrix, number.cex = .7, tl.pos = "d", tl.cex = 0.6)

```

***
## Standardization

In standardization means of all variables are in zero. That is, variables have distributed around zero. This can be seen in summary table (compare with original summary above).

Variable crime rate has been changed to categorical variable with 4 levels: low, med_low, med_high and high. Each class includes quantile of data (25%). 

Train and test sets have been created by dividing original (standardized) data to two groups randomly. 80% belongs to train set and 20% to test set.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Standardization of the dataset
boston_scaled <- scale(Boston)
summary(boston_scaled)
# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
# create a quantile vector of crim
bins <- quantile(boston_scaled$crim)
# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))
# removing original crim
boston_scaled <- dplyr::select(boston_scaled, -crim)
# adding the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# Creation of train and test data sets
n <- nrow(boston_scaled)
# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
# create train set
train <- boston_scaled[ind,]
# create test set 
test <- boston_scaled[-ind,]
# save the correct classes from test data
correct_classes <- test$crime
# remove the crime variable from test data
test <- dplyr::select(test, -crime)


```

Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot.
  

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Fit the linear discriminant analysis on the train set.
# Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables.
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)
# Draw the LDA (bi)plot.
# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 2)

```

 Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results. 


```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# Cross tabulate the results with the crime categories from the test set. Comment on the results. (0-3 points)
# cross tabulate the results
tab <- table(correct = correct_classes, predicted = lda.pred$class) %>% addmargins()
tab
correct_pred <- tab[1,1] + tab[2,2] + tab[3,3] + tab[4,4]
incorrect_pred <- tab[5,5] - (tab[1,1] + tab[2,2] + tab[3,3] + tab[4,4]) 
pred_error <- incorrect_pred / tab[5,5]
pred_error
```

Calculate the distances between the observations. Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results. (0-4 points)  

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# center and standardize variables

data(Boston)

# Reload the Boston dataset and standardize the dataset.
boston_scaled <- scale(Boston)
# class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)

# Calculate the distances between the observations.
# euclidean distance matrix
dist_eu <- dist(boston_scaled)
# look at the summary of the distances
# summary(dist_eu)
# manhattan distance matrix
# dist_man <- dist(Boston, method = 'manhattan')
# look at the summary of the distances
# summary(dist_man)

# Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results.
# k-means clustering. Centers = number of clusters
km <-kmeans(boston_scaled, centers = 3)
# plot the Boston dataset with clusters
# pairs(Boston[6:10], col = km$cluster)


# Investigate what is the optimal number of clusters
set.seed(123)
# determine the number of clusters
k_max <- 10
# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})
# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(boston_scaled, centers = 2)
# plot the Boston dataset with clusters
# pairs(boston_scaled, col = km$cluster)
ggpairs(boston_scaled, mapping = aes(col = as.factor(km$cluster), alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

```

Bonus: Perform k-means on the original Boston data with some reasonable number of clusters (> 2). Remember to standardize the dataset. Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model. Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influencial linear separators for the clusters? 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

Super-Bonus: 

Adjust the code: add argument color as a argument in the plot_ly() function. Set the color to be the crime classes of the train set. Draw another 3D plot where the color is defined by the clusters of the k-means. How do the plots differ? Are there any similarities? 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color='classes')

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color='km$cluster')

```




```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```




```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```




```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```











