---
title: "Analysis of longitudinal and repeated measures data"
author: "Jussi Jaatinen"
output: html_document
---

# Analysis of longitudinal and repeated measures data

If the same person makes several observations in same experiment or the weight of the same animal is measured several times during the longer time or in different conditions, those measurements are hardly independent. This must be also taken into account in analysis models of such kind of data. 

Two such kind of datasets are wrangled and analyzed in this excercise. First set, RATS, includes information about different nutritions and their influence to the weight of rats during 9 weeks period. In the second set, BPRS, 40 male subjects were randomly assigned to two groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment and then at weekly intervals for eight weeks.The scale is used to evaluate patients suspected of having schizophrenia.

Both datasets have been converted to long form for analyses. The data wrangling code can be explored here.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)

# read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
# Factor variables ID and Group
RATS$ID <- as.factor(RATS$ID)
RATS$Group <-as.factor(RATS$Group)
# Glimpse the data
glimpse(RATS)
# Convert data to long form
RATSL <- RATS %>% gather(key = WD, value = Weight, -ID, -Group)
RATSL <- RATSL %>% mutate(Time = as.integer(substr(RATSL$WD, 3, 4)))  
# Glimpse the data
glimpse(RATSL)

# Plot the RATSL data
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line() + aes(linetype = Group) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) + theme(legend.position = "top")
  
# create a regression model RATS_reg
RATS_reg <- lm(RATSL$Weight ~ RATSL$Time + RATSL$Group, data = RATSL)

# print out a summary of the model
summary(RATS_reg)

# Create a random intercept model
RATS_ref <- lmer(Weight ~ Time + Group + (1 | ID), data = RATSL, REML = FALSE)

# Print the summary of the model

summary(RATS_ref)

# create a random intercept and random slope model
RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref1)

# perform an ANOVA test on the two models
anova(RATS_ref1, RATS_ref)

# create a random intercept and random slope model with the interaction
RATS_ref2 <- RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID) + (Time | Group), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref2)

# perform an ANOVA test on the two models
anova(RATS_ref2, RATS_ref1)

# draw the plot of RATSL with the observed Weight values
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(RATS_ref2)

# Create a new column fitted to RATSL
RATSL <- mutate(RATSL, Fitted = Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)
library(ggplot2)

# Read the BPRS data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
# Look at the (column) names of BPRS
names(BPRS)
# Look at the structure of BPRS
str(BPRS)
# Print out summaries of the variables
summary(BPRS)
# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
# Convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(BPRSL$weeks, 5, 5)))
# Take a glimpse at the BPRSL data
glimpse(BPRSL)

# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
  
# Standardise the variable bprs
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = scale(bprs)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL)

# Plot again with the standardised bprs
ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")

# Number of weeks, baseline (week 0) included
n <- BPRSL$week %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise( mean = mean(bprs), se = sd(bprs) / sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSS)

# Plot the mean profiles
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
BPRSL8S1 <- BPRSL %>%
  filter(week > 0, bprs < 45) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()
ggplot(BPRSL8S1, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
  
# Perform a two-sample t-test
t.test(mean ~ treatment, data = BPRSL8S1, var.equal = TRUE)

# Add the baseline from the original data as a new variable to the summary data
BPRSL8S2 <- BPRSL8S %>%
  mutate(baseline = BPRS$week0)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + treatment, data = BPRSL8S2)

# Compute the analysis of variance table for the fitted model with anova()

anova(fit)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

```








During the Analysis exercise you will explore both data sets graphically and perform some initial analyses that require various data conversions. Finally you will create and test a few advanced statistical models and, of course, interpret the results in all phases of the analysis.
The Analysis exercise focuses on exploring your data and performing and interpreting simple straightforward methods (based on summary measures of data) and more advanced methods (called linear mixed effects models). For completing the Analysis exercises, include all the codes, your interpretations and explanations in the R Markdown file chapter6.Rmd. 

Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points)

Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points)
 





