---
title: "Analysis of longitudinal and repeated measures data"
author: "Jussi Jaatinen"
output: html_document
---

# Analysis of longitudinal and repeated measures data

If the same person makes several observations in same experiment or the weight of the same animal is measured several times during the longer time or in different conditions, those measurements are hardly independent. This must be also taken into account in analysis models of such kind of data. 

Two such kind of datasets are wrangled and analyzed in this excercise. First set, RATS, includes information about different nutritions and their influence to the weight of rats during 9 weeks period. In the second set, BPRS, 40 male subjects were randomly assigned to two groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment and then at weekly intervals for eight weeks.The scale is used to evaluate patients suspected of having schizophrenia.

Both datasets have been converted to long form for analyses. 

## RATS dataframe

Here is the original RATS dataframe and same data wrangled to long form with class changes. Variables ID and Group have been changed to factor type.
As seen in tables, structure of dataframes are quite different, although they include the same information. A new variable "time" has been added to long form dataframe, RATSL. It is extracted from variable "WD" by picking only numbers and changing class to integer.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)

# read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# Glimpse the data
glimpse(RATS)

# Factor variables ID and Group
RATS$ID <- as.factor(RATS$ID)
RATS$Group <-as.factor(RATS$Group)

# Convert data to long form
RATSL <- RATS %>% gather(key = WD, value = Weight, -ID, -Group)
RATSL <- RATSL %>% mutate(Time = as.integer(substr(RATSL$WD, 3, 4)))  
# Glimpse the data
glimpse(RATSL)

```
### Plot of the RATSL data

Here each individual rat has separate curves and three groups have marked with different line types.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)


# Plot the RATSL data
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line() + aes(linetype = Group) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) + theme(legend.position = "top")

```

#### Normal linear regression model (not for repeated measures)

This is a normal linear model analysis without any interactions or random variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# create a regression model RATS_reg
RATS_reg <- lm(RATSL$Weight ~ RATSL$Time + RATSL$Group, data = RATSL)

# print out a summary of the model
summary(RATS_reg)
```

#### Random intercept model

Here a random variable intercept (1 | ID) has been added and Linear Mixed Effects Regression model analysis (lmer function) has been executed. This means that individual differences in intercepts between subjects have been taken into account as an random variable. In results we can see that all fixed effects are significant.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Create a random intercept model
RATS_ref <- lmer(Weight ~ Time + Group + (1 | ID), data = RATSL, REML = FALSE)

# Print the summary of the model

summary(RATS_ref)
```

#### Random slope model

Here is another model with random slope model (Time | ID).

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# create a random intercept and random slope model
RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref1)
```

#### ANOVA of "Random  slope model" and "Random intercept model"

Here we can see results of ANOVA for two aforementioned models. Here we can see that due to Chi^2^ test, there is a significant difference between these two models.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# perform an ANOVA test on the two models
anova(RATS_ref1, RATS_ref)
```

#### Random slope model with the interaction

And here is another random variable (Time | Group) added.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# create a random intercept and random slope model with the interaction

RATS_ref2 <- lmer(Weight ~ Time + Group + (Time | ID) + (Time | Group), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref2)

```

#### ANOVA of "Random slope model with interaction" and "Random slope model"

Here again ANOVA for two latest models.  Due to Chi^2^ test, there is no significant difference between these two models.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# perform an ANOVA test on the two models
anova(RATS_ref2, RATS_ref1)

```

#### Plot of RATSL with the observed Weight values

Comparison of observed and fitted weight values. 

Observed values

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# draw the plot of RATSL with the observed Weight values
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")
```

#### Plot of RATSL with the Fitted values of weight

Fitted values

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Create a vector of the fitted values
Fitted <- fitted(RATS_ref2)

# Create a new column fitted to RATSL
RATSL <- mutate(RATSL, Fitted = Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")

```

## BPRS dataframe

As well as in RATS dataframe same data wrangling to long form with class changes has been made. Variables "treatment" and "subject" have been changed to factor type.
As seen in tables, structure of dataframes are quite different, although they include the same information. Also here, a new integer variable "week" has been added to long form dataframe, BPRSL. 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)
library(ggplot2)

# Read the BPRS data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)

glimpse(BPRS)

# Look at the (column) names of BPRS
# names(BPRS)
# Look at the structure of BPRS
# str(BPRS)
# Print out summaries of the variables
# summary(BPRS)
# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
# Convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(BPRSL$weeks, 5, 5)))
# Take a glimpse at the BPRSL data
glimpse(BPRSL)

# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

#### Standardization of the variable "bprs"

Here, variable "bprs" has been standardized and the new variable "stdbprs" has been created and added to BPRSL dataframe.
Men who have higher BPRS values at the beginning seem to have higher values throughout the study. 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Standardise the variable bprs
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = scale(bprs)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL)

# Plot again with the standardised bprs
ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")
```

#### Means and standard errors

Means (over all subjects) and standard errors have been calculated. In the figure below differences between treatments can be seen. Due to confidental intervals are overlapping, probability to statistically significant difference is weak.
  
```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Number of weeks, baseline (week 0) included
n <- BPRSL$week %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise( mean = mean(bprs), se = sd(bprs) / sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSS)

# Plot the mean profiles
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")
```

#### Boxplot of the mean versus treatment

Here is boxplot of both treatment groups.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
```

#### Outlier

Here we can see in the boxplot above an outlier, which may be an error. It also moves mean of the treatment 2 upwards. It can be removed by filtering out bprs values above 60. Below is the new boxplot with outlier removed. Mean (white square) is now closer to median (black horizontal line).

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
BPRSL8S1 <- BPRSL %>%
  filter(week > 0, bprs < 60) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()
ggplot(BPRSL8S1, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
  
```

#### T-test

For testing difference between means of treatments, common two sample t-test is suitable. According to it, p value 0.75 is far from significance. That is, treatments do not have statistically significant difference.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Perform a two-sample t-test
t.test(mean ~ treatment, data = BPRSL8S1, var.equal = TRUE)
```

#### Adding baseline 

Here we add the baseline from the original data as a new variable to the summary data.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Add the baseline from the original data as a new variable to the summary data
BPRSL8S2 <- BPRSL8S %>%
  mutate(baseline = BPRS$week0)
glimpse(BPRSL8S2)
```

#### Linear model with baseline and treatment

Here we run lm -function "mean" as target and "baseline" and "treatment" as explanatory variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + treatment, data = BPRSL8S2)
fit

```

#### Anova of linear model

Even with baseline corrected, treatments have not significant difference (p = 0.81).

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Compute the analysis of variance table for the fitted model with anova()

anova(fit)
```
















