---
title: "Analysis of Boston data"
author: "Jussi Jaatinen"
output: html_document
fig_caption: true
fig_width: 15
fig_height: 8
code_folding: hide
---

# Analysis of Boston data

### Overview of data

Boston data is included in R-package as a demonstartion or example.

Dataset contains social, environmental and economical information about great Boston area. It includes following variables:

- crim = per capita crime rate by town             
- zn = proportion of residential land zoned for lots over 25,000 sq.ft.    
- indus =  proportion of non-retail business acres per town				
- chas = Charles River dummy variable 
- nox =  nitrogen oxides concentration (parts per 10 million)
- rm = average number of rooms per dwelling
- age  = proportion of owner-occupied units built prior to 1940
- dis = weighted mean of distances to five Boston employment centres
- rad = index of accessibility to radial highways
- tax =  full-value property-tax rate per $10000  
- ptratio = pupil-teacher ratio by town
- black = 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
- lstat = lower status of the population (percent)
- medv = median value of owner-occupied homes in $1000s

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Loading necessary libraries
library(MASS)
library(dplyr)
library(plotly)
library(corrplot)
library(ggplot2)
library(GGally)
# load the data
data("Boston")
```

### Structure and the dimensions of the data 

Dataset has 14 variables and 506 observations and all variables are numerical.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
str(Boston)
dim(Boston)

```

### Summary, graphical presentation of data and correlations

As seen in pairs plot, most of the variables are not normally distributed. Most of them are skewed and some of them are bimodal. Correlations between variables are better viewed in correlation plotting, where on the upper-right side the biggest circles indicate highest correlations (blue = positive or red = negative). Corresponding number values are mirrored on the lower-left side.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
summary(Boston)
# pairs(Boston)
# ggpairs(Boston, mapping = aes(), lower = list(combo = wrap("facethist", bins = 40)), upper = list(continuous = wrap("cor", size = 12)))
ggpairs(Boston, mapping = aes(corSize = 8), lower = list(combo = wrap("facethist", bins = 100)))

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits = 2)
# visualize the correlation matrix
# corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
corrplot.mixed(cor_matrix, number.cex = .7, tl.pos = "d", tl.cex = 0.6)

```

***

### Standardization

In standardization means of all variables are in zero. That is, variables have distributed around zero. This can be seen in summary table (compare with original summary above).

Variable crime rate has been changed to categorical variable with 4 levels: low, med_low, med_high and high. Each class includes quantile of data (25%). 

Train and test sets have been created by dividing original (standardized) data to two groups randomly. 80% belongs to train set and 20% to test set.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Standardization of the dataset
boston_scaled <- scale(Boston)
summary(boston_scaled)
# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
# create a quantile vector of crim
bins <- quantile(boston_scaled$crim)
# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))
# removing original crim
boston_scaled <- dplyr::select(boston_scaled, -crim)
# adding the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# Creation of train and test data sets
n <- nrow(boston_scaled)
# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
# create train set
train <- boston_scaled[ind,]
# create test set 
test <- boston_scaled[-ind,]
# save the correct classes from test data
correct_classes <- test$crime
# remove the crime variable from test data
test <- dplyr::select(test, -crime)


```

### Linear discriminant analysis (LDA)

In linear discriminant analysis (LDA) only the train set (80% of data) has been analysed. Target variable is the new categorical variable, crime rate (low, med_low, med_high, high). In LDA model all other variables of the data set are used as predictor variables (see Overview of data).

In biplot below can be seen that variable "rad" (index of accessibility to radial highways) has extremely high influence to LD1 and LD2 if compared to the other variables. In biplot all horizontal vectors describes contribution to LD1 dimension (x-axis) and vertical vectors LD2-dimension (y-axis). Sign of coefficient of linear discriminant determines the direction of vector. The longer the vector, the bigger is influence. Most of the vectors contribute both LD1 and LD2. Because in biplot two dimensions are illustrated, directions of most of variables are in different angles between LD1 and LD 2. For example, in the LDA table below the most significant variable of LD1 "rad" has coefficients LD1 = 3.27 and LD2 = 1.05. They are directly readable as coordinates of the arrow head. Similarly the second most significant variable of LD2, "nox" has its head ccordinates in (-0.69, 0.29). LDA1 alone explains 0.95% of model. LD2 explains 3% and LD3 only 1%.
  

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Fit the linear discriminant analysis on the train set.
# Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables.
lda.fit <- lda(crime ~ ., data = train)
# print the lda.fit object
lda.fit
# Linear discriminant analysis on the train set.
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)
# Draw the LDA (bi)plot.
# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

```

### Predictive power of the model

In the test dataset catecorigal crime variable has been removed. In the table below true values of the original test data and predicted values of the test data (crime removed) are cross-tabulated. Total amount of observations is 102 (506/5 +1).
In the table on diagonal axis (from top-left corner) are true values (sum = 76) and all others are predicted values (sum = 26). Prediction error is 26/102 ≈ 0.25

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# Cross tabulate the results with the crime categories from the test set. Comment on the results. (0-3 points)
# cross tabulate the results
tab <- table(correct = correct_classes, predicted = lda.pred$class) %>% addmargins()
tab
correct_pred <- tab[1,1] + tab[2,2] + tab[3,3] + tab[4,4]
#correct_pred
incorrect_pred <- tab[5,5] - (tab[1,1] + tab[2,2] + tab[3,3] + tab[4,4]) 
#incorrect_pred
pred_error <- incorrect_pred / tab[5,5]
#pred_error
```

### Calculation of distances between the observations and optimal number of clusters

In this model euclidean distance matrix has been calculated.  Results can be seen in table below. 
By using K-means algorithm, the optimal number of clusters can be investigated. When TWSS (total within sum of squares) drops significally, it indicates optimal number of clusters. In this case optimal number of clusters is 2 or 3. In the first plotting, data has classified into two and in the second plotting three clusters.


```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# center and standardize variables

data(Boston)

# Reload the Boston dataset and standardize the dataset.
boston_scaled <- scale(Boston)
# class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)

# Calculation of the distances between the observations using euclidean method
dist_eu <- dist(boston_scaled)
# look at the summary of the distances
summary(dist_eu)

# k-means clustering. Centers = number of clusters
# km <-kmeans(boston_scaled, centers = 3)
# plot the Boston dataset with clusters

# Investigate what is the optimal number of clusters
set.seed(123)
# determine the number of clusters
k_max <- 8
# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})
# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line', xlim = c(1, k_max), xlab = "Number of clusters", ylab = "Total Withis Sum of Squares")

# k-means clustering
km <-kmeans(boston_scaled, centers = 2)
# plot the Boston dataset with clusters
# pairs(boston_scaled, col = km$cluster)
ggpairs(boston_scaled, mapping = aes(col = as.factor(km$cluster), alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20), upper = NULL))

km <-kmeans(boston_scaled, centers = 3)
# plot the Boston dataset with clusters
# pairs(boston_scaled, col = km$cluster)
ggpairs(boston_scaled, mapping = aes(col = as.factor(km$cluster), alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20), upper = NULL))

```

### Bonus 

Here LDA is calculated with the clusters as target classes. All other variables in the Boston data are predictor variables. In LDA tables and biplots, differences between number of clusters can be seen. Variable "rad" is the most influencial linear separator for the clusters in LD1 and variable "zn" in LD2. At the moment Knitting does not accept my code. Code is visible below. I tried to fix this later.


data("Boston")
boston_scaled <- scale(Boston)
boston_scaled <- as.data.frame(boston_scaled)

the function for lda biplot arrows

lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

km3 <-kmeans(boston_scaled, centers = 3)

km4 <-kmeans(boston_scaled, centers = 4)

km5 <-kmeans(boston_scaled, centers = 5)

km6 <-kmeans(boston_scaled, centers = 6)

clu3 <- as.factor(km3$cluster)

clu4 <- as.factor(km4$cluster)

clu5 <- as.factor(km5$cluster)

clu6 <- as.factor(km6$cluster)

lda.fit3 <- lda(clu3 ~ ., data = boston_scaled)

lda.fit3

lda.fit4 <- lda(clu4 ~ ., data = boston_scaled)

lda.fit4

lda.fit5 <- lda(clu5 ~ ., data = boston_scaled)

lda.fit5

lda.fit6 <- lda(clu6 ~ ., data = boston_scaled)

lda.fit6

target classes as numeric

classes <- as.numeric(clu3)

plot(lda.fit3, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

classes <- as.numeric(clu4)
plot(lda.fit4, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

classes <- as.numeric(clu5)
plot(lda.fit5, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

classes <- as.numeric(clu6)
plot(lda.fit6, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)


### Super-Bonus

Adjust the code: add argument color as a argument in the plot_ly() function. Set the color to be the crime classes of the train set. Draw another 3D plot where the color is defined by the clusters of the k-means. How do the plots differ? Are there any similarities? 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

data("Boston")
boston_scaled <- scale(Boston)
summary(boston_scaled)
# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
# create a quantile vector of crim
bins <- quantile(boston_scaled$crim)
# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))
# removing original crim
boston_scaled <- dplyr::select(boston_scaled, -crim)
# adding the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables.
lda.fit <- lda(crime ~ ., data = train)
# print the lda.fit object
# lda.fit

model_predictors <- dplyr::select(train, -crime)
# check the dimensions
# dim(model_predictors)
# dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
#Next, install and access the plotly package. Create a 3D plot (Cool!) of the columns of the matrix product by typing the code below.

km8 <-kmeans(model_predictors, centers = 4)
clu8 <- as.factor(km8$cluster)

crime_colour <- as.numeric(train$crime)
k_mean_colour <- as.numeric(clu8)

```

##### Train data classified by Crime (1 = low, 4 = high)

Data points are of course in same positions. Grouping differs slightly in main group if colours are coded either by crime or by cluster. In the separate group high-crime is well isolated whereas in clusters, there are two of them. If colours are coded by crime, particularly the high-crime is better gathered to one group.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=crime_colour)
```

##### Train data classified by Clusters 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color=k_mean_colour)
 
   
```










