---
title: "Analysis of Alcohol consumption in two Portugese secondary schools"
author: "Jussi Jaatinen"
output: html_document
---

# Analysis of Alcohol consumption in two Portugese secondary schools


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(boot)

alc <- read.table("~/Documents/GitHub/IODS-project/data/alc.txt")
```

***

### Data, hyphotheses and selected explanatory variables

The data of this study has been collected from two Portugese second grade schools by using school reports and questionnaires. Aim of this analysis is to find relationships between high alcohol consumption and hypothesized explanatory variables. 

Hyphotheses: 

- Males consume more alcohol  
- It is more common to use alcohol with friends outside home
- Good family relations decrease drinking 
- Drinking may cause absences

According to hyphotheses, the selected explanatory variables are sex, go out, family relations and absences. According to first hyphothesis, all analyses have been divided to two groups, females and males.

In dataset there are lot of background variables like demographic, social relationships, parents´education and employment, or living area as well as grades, absences or health. Full descriptions of study is available here: https://archive.ics.uci.edu/ml/datasets/Student+Performance. For alcohol consumption related study, the original data has been modified by combining and averaging weekend and weekdays alcohol consumptions. As result two new variables have been created: alc_use (Likert, 1-5, where 1 = very low, 5 = very high) and high_use (logical, TRUE > 2). Total number of variables after adding is 35. Number of observations is 382, which 198 females, 184 males.

***

### All variables and distributions

```{r echo=FALSE, message=FALSE, warning=FALSE}

gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```

***

### Distributions and observations of chosen variables

##### Relationship between alcohol consumption and sex

As seen in table below, in generally, males consume more alcohol. In low consumption group (very low or low = FALSE), 58% were females and 42% males, whereas in higher consumption group only 37% were females and 63% males.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

table(alc$sex, alc$high_use) %>% addmargins() 

```

##### Relationship between alcohol consumption and going out with friends

Having more freetime outside home with friends increases alcohol consumption in both sexes significantly.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# initialize a plot of high_use and going out with friends
g1 <- ggplot(alc, aes(x = high_use, y = goout, col = sex))
g1 + geom_boxplot() + xlab("Alcohol consumption low / high") + ylab("Frequency of going out with friends")
```

##### Relationship between alcohol consumption and absences

Although absences increase due to higher alcohol consumption, difference is quite moderate, thus larger in males.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# initialise a plot of high_use and absences
g2 <- ggplot(alc, aes(x = high_use, y = absences, col = sex))
g2 + geom_boxplot() + xlab("Alcohol consumption low / high") + ylab("Absences") 

```

##### Relationship between alcohol consumption and family relations

Although same tendency occur also in this case, there also might be two-sided correlation between alcohol consumption and family relations. Which of them is more cause or consequence is unclear. Bad relationships in family may lead bigger alcohol consumption, but also drinking may evoke collisions between family members.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# initialise a plot of high_use and family relations
g3 <- ggplot(alc, aes(x = high_use, y = famrel, col = sex))
g3 + geom_boxplot() + xlab("Alcohol consumption low / high") + ylab("Family relations")

```

***

### Logistic regression 

All chosen variables are statistically significant. Especially sex is most significant of all and its odd ratio was the highest (2.84). This means that propability for high alcohol consumption in males is almost three times higher than in females. Another strongly influencing variable is "Going out with friends" (goout). None of the included variables include 1 in confidental interval, which means they all have meaningful influence to the model. 

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# Logistic regression model with chosen variables
m <- glm(high_use ~ famrel + absences + sex + goout, data = alc, family = "binomial")
# Summary of the model
summary(m)
# Coefficients of the model
# coef(m)
# Odds ratios (OR)
OR <- coef(m) %>% exp
# Confidence intervals (CI)
CI <- confint(m) %>% exp
# Oodds ratios with confidence intervals
cbind(OR, CI)
```

***

### Predictive power of the model

With chosen variables, predictive power of model (penalty) is ~ 0.20 (TRUE-FALSE + FALSE-TRUE pairs summed). This means that about 4/5 of predictions are correct (TRUE-TRUE + FALSE-FALSE pairs summed).

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}
# predict the probability of high_use
probabilities <- predict(m, type = "response")
# Adding the new variable, predicted probabilities, to alc
alc <- mutate(alc, probability = probabilities)
# Define "high use" in probabilities
alc <- mutate(alc, prediction = (probability > 0.5))
# Tabulate high_use versus the predictions with propabilities and margin
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table() %>% addmargins()
# Plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction)) + geom_point() + aes(col = prediction)
g 
```

##### Mean prediction error

Mean prediction error of the model.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# Define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}
# Compute the average number of wrong predictions in the (training) data by calling loss function
loss_func(class = alc$high_use, prob = alc$probability)

# compute the average number of wrong predictions in the (training) data
```

***

### K-fold cross-validation 

K-fold cross-validation gives slighly higher prediction error.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

cvk <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = nrow(alc))
cvk$delta[1]

```

***

### 10-fold cross-validation on my model. 

This model have smaller prediction error (0.20) using 10-fold cross-validation compared to the model introduced in DataCamp (which had about 0.26 error). This means that chosen variables in my model predict better high alcohol consumption.

```{r echo=FALSE, message=FALSE, warning=FALSE, comment=''}

# 10-fold cross-validation 
cv10 <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)
cv10$delta[1]

```

***

### Cross-validation to compare the performance of different logistic regression models

The highest amount of predictors is in my original model introduced earlier. Other models are reduced combinations of original model or added to the model previously unused variables (N=18). In 10-fold cross validation, the smallest penalty loss (0.20) was in my original model. Changing variables increased or decreased penalty loss depending on significancy of chosen variables.

```{r echo=FALSE, message=FALSE, comment=''}

# find the model with glm()
m1 <- glm(high_use ~ famrel + absences + sex + goout, data = alc, family = "binomial")
m2 <- glm(high_use ~ famrel + absences + sex, data = alc, family = "binomial")
m3 <- glm(high_use ~ famrel + absences, data = alc, family = "binomial")
m4 <- glm(high_use ~ famrel, data = alc, family = "binomial")
m5 <- glm(high_use ~ absences + sex + goout, data = alc, family = "binomial")
m6 <- glm(high_use ~ famrel + absences+ goout, data = alc, family = "binomial")
m7 <- glm(high_use ~ famrel + absences + sex, data = alc, family = "binomial")
m8 <- glm(high_use ~ absences, data = alc, family = "binomial")
m9 <- glm(high_use ~ absences + sex + goout, data = alc, family = "binomial")
m10 <- glm(high_use ~ famrel + sex + goout, data = alc, family = "binomial")
m11 <- glm(high_use ~ famrel + absences + sex, data = alc, family = "binomial")
m12 <- glm(high_use ~ sex, data = alc, family = "binomial")
m13 <- glm(high_use ~ famrel + sex, data = alc, family = "binomial")
m14 <- glm(high_use ~ absences + sex, data = alc, family = "binomial")
m15 <- glm(high_use ~ sex + goout, data = alc, family = "binomial")
m16 <- glm(high_use ~ famrel + goout, data = alc, family = "binomial")
m17 <- glm(high_use ~ absences + goout, data = alc, family = "binomial")
m18 <- glm(high_use ~ goout, data = alc, family = "binomial")
m19 <- glm(high_use ~ famrel + absences + sex + G3 + freetime + age + failures + famsize + romantic + health + goout, data = alc, family = "binomial")


model_vector = c("m1", "m2", "m3", "m4", "m5", "m6", "m7", "m8", "m9", "m10", "m11", "m12", "m13", "m14", "m15", "m16", "m17", "m18", "m19")
# print out a summary of the model
for (i in model_vector) {
sum <- summary(get(i))
coe <- coef(get(i))
#print(sum)
#print(coe)
OR <- coe %>% exp
CI <- confint(get(i)) %>% exp
ORCI <- cbind(OR, CI)
# print(OR)
print(ORCI)
# cv <- cv.glm(data = alc, cost = loss_func, glmfit = get(i), K = nrow(alc))
cv <- cv.glm(data = alc, cost = loss_func, glmfit = get(i), K = 10)
print(cv$delta[1])


probabilities <- predict(get(i), type = "response")
alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prediction = (probability > 0.5))
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table() %>% addmargins()
# g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction)) + geom_point() + aes(col = prediction)
# + geom_point() + aes(col = prediction)

loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

lf <- loss_func(class = alc$high_use, prob = alc$probability)
lf
}

```

***